Forcing for Isabelle

### 2018-04-12 blueprint.tex
prueba de separación, primera parte

\renewcommand{\PP}{\mathbb{P}}
\renewcommand{\app}{\mathrm{App}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\newcommand{\forceisa}{\mathtt{forces}}
\newcommand{\equ}{\mathbf{e}}
\newcommand{\bel}{\mathbf{b}}
\newcommand{\atr}{\mathit{atr}}
\newcommand{\reee}{\mathrel{\mathtt{eee}}}
\DeclareMathOperator{\val}{val}

@@

\begin{definition}
  $\tau$ is a \emph{$\PP$-name} if its elements are of the form
  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$.
\end{definition}
The usual way to define a property by well-founded recursion is to
define the ``characteristic function'' of that property. In this case,
define by recursion the function $F:M\to\om$ such that $F(x)=1$  if
$x$ is a $\PP$-name, or $F(x)=0$ otherwise.

@@

>>>>

\renewcommand{\PP}{\mathbb{P}}
\renewcommand{\app}{\mathrm{App}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\newcommand{\forceisa}{\mathtt{forces}}
\newcommand{\equ}{\mathbf{e}}
\newcommand{\bel}{\mathbf{b}}
\newcommand{\atr}{\mathit{atr}}
\newcommand{\concat}{\mathbin{@}}
\newcommand{\reee}{\mathrel{\mathtt{eee}}}
\DeclareMathOperator{\val}{val}

@@

\begin{definition}
  $\tau$ is a \emph{$\PP$-name} if its elements are of the form
  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$-name.
\end{definition}
The usual way to define a property by well-founded recursion is to
define the ``characteristic function'' of that property. In this case,
define by recursion the function $F:M\to\om$ such that $F(x)=1$  if
$x$ is a $\PP$-name, or $F(x)=0$ otherwise.

@@

\section{The axioms in the generic extension}
\subsection{The axiom of Separation}
In this subsection we will show that the Separation Axiom scheme holds
in the generic extension $M[G]$.

\begin{lemma}
  Let $G$ be $M$-generic for $\PP$. For every $\phi\in\formula$ with
  three free variables, the $\phi$-instance of Separation holds in
  $M[G]$.
\end{lemma}
\begin{proof}
  Let $G$ as in the hypothesis, $\phi(x,z,v)$ be in \formula{} and let
  $c,w\in M[G]$. Following Kunen \cite[Lemma IV.2.6]{kunen2011set},
  we have to show that 
  \[
  S\defi\{x\in c : M[G]\models \phi(x,c,w)\} \in M[G].
  \]
  From this, using absoluteness, we will be able to derive
  $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
  $\val(\pi)=c$ and $\val(\sig)=w$.
  We will show that the name
  \[
  \tau \defi \{\lb\th,p\rb: \th\in\dom(\pi) \y p\in \PP\y
  M\models p\forces(\th\in\pi\y\phi(\th,\pi,\sig)\}
  \]
  belongs to $M$; for this we will use that an appropriate instance of
  Separation holds in $M$. Later we will see that $\tau$ names $S$.

  First, we slightly rewrite $\tau$:
  \[
  \tau \defi \{w\in\dom(\pi)\times\PP :\exists \th,p (w=\lb\th,p\rb \y 
  M\models p \forces(\th\in\pi\y\phi(\th,\pi,\sig))\}
  \]

  The formula giving the relevant instance of Separation in $M$ will be the
  following:
  \[
  \psi \defi \exists \th,p : x=\lb\th,p\rb \y \forceisa(\PP,\leq,\1,p,\th\in\pi\y\phi(\th,\pi,\sig)),
  \]
  where $\PP,\leq,\1,\pi,\sig$ are parameters.
  \begin{framed}
    I'm changing the order of arguments of $\forceisa$. Also, we
    will need to show that, in $M$, a version of separation for formulas
    with  $5$ free variables holds.

    Also, we need to show that $\dom(\pi)\times\PP\in M$.
  \end{framed}
  Writing this in de Bruijn notation, gives us something like this:
  \[
  \psi \defi  \exists\,\exists\,\verb|pair_fm|(0,1,2) \y
  \forceisa(4,5,6,7,0\in 1\y\phi(0,1,3)).
  \]
  From $M\models \mathtt{Separation}(\psi)$ we have
  \begin{multline*}
    \forall u,l,P,s,r\in M \forall d\in M:\\ 
    M,[d,r,s,P,l,u] \models\exists y \forall x
    (x\in y \lsii x\in d \y \psi(P,l,u,x,r)).
  \end{multline*}
  We instantiate the universal quantifiers using the given parameters
  and the domain $\dom(\pi)\times\PP$, while changing the free variables in
  the formula with de Bruijn indices corresponding to the order in the
  environment.
  \[
  M,[\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models\exists y \forall x
  (x\in y \lsii x\in 0 \y \psi(3,4,5,x,1)).
  \]
  Extracting the two outermost quantifiers and shifting indices, we get
  \begin{multline*}
    \exists y\in M \forall x\in M: \\
    M,[x,y,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
    0\in 1 \lsii 0\in 2 \y \psi(5,6,7,0,3).
  \end{multline*}
  Let $\tau\in M$ witness the $\exists y$. So, by absoluteness,
  \begin{multline*}
    \forall x\in M: 
    x\in\tau \lsii  x\in\dom(\pi)\times\PP \y{}\\ 
  M,[x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
  \psi(5,6,7,0,3).
  \end{multline*}
  Let $\mathit{env}\defi
  [x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]$. By transitivity
  of $M$ we can dispense of the bound on $x$, and we get
  \[
  \tau = \{x\in\dom(\pi)\times\PP :
  M, \mathit{env}\models
  \psi(5,6,7,0,3)\},
  \]
  i.e., the relevant name is in $M$.

  Now we turn to the proof that $\val(\tau)=S$.

  Let's first focus into the predicate defining $\tau$ by separation. By expanding $\psi$ we have
  \[
   M, \mathit{env}\models
   \exists \th,p : 0=\lb\th,p\rb \y
   \forceisa(5,6,7,p,\th\in 3\y\phi(\th,3,4)).
  \]
  We proceed as before, using absoluteness and replacing variables by
  corresponding indices.
  \[
  \exists \th,p : x =\lb\th,p\rb \y  M, [\th,p]\concat\mathit{env}\models
  \forceisa(7,8,9,1,0\in 5\y\phi(0,5,6)).
  \]

  Considering that only a fraction of elements
  of the environment  are necessary to interpret
  $\forceisa(7,8,9,1,0\in\pi\y\phi(0,5,6))$ and by permuting, we can
  replace
  \[
  M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
  \forceisa(7,8,9,1,0\in 5 \y\phi(0,5,6))
  \]
  by
  \[
  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
  \forceisa(0,1,2,3,4\in 5 \y\phi(4,5,6))
  \]
  By the Truth Lemma, this is equivalent to
  \[
  \forall F\ \PP\text{-generic}, p\in F \limp
  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
  0\in1  \y\phi(0,1,2).
  \]

  Now we calculate $\val(\tau)$.
  \begin{multline*}
    \val(\tau) = \\
    \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
    \y  \\
    \forall F\ \PP\text{-generic}, p\in F \limp
    M[F],[\val(\th),\val(\pi),\val(\sig)]\models
    0\in 1 \y\phi(0,1,2)\}\bigr).
  \end{multline*}
  Hence
  \begin{multline*}
    \val(\tau) = 
    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
    \y \forall F\ \PP\text{-generic}, \\
   p\in F \longrightarrow
  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
  0\in 1 \y\phi(0,1,2)\}.
  \end{multline*}
  This set includes
  \begin{multline*}
    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
    \y  \\
   p\in G \longrightarrow
   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
   0\in 1 \y\phi(0,1,2)\},
  \end{multline*}
  that is
  \begin{multline*}
    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
    \y 
   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
   0\in 1 \y\phi(0,1,2)\}
  \end{multline*}
\end{proof}
